---
title: "Assignment 2"
subtitle: 'Cultural Data Science'
author: "Glorija Stvol"
date: "November 6, 2024"
output: pdf_document
---

## Preparing the data

Load  the 'divorce_margarine' dataset  from  the 'dslabs'package.
Load the ‘GSSvocab’dataset from the ‘car’package.

```{r, include=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org")) #included to knit properly

#install packages
install.packages("dslabs")
install.packages("car")
```

```{r, include=FALSE}
#load libraries
library(dslabs)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(car)
```

```{r}
#load the divorce_margarine dataset 
data("divorce_margarine")
df <- divorce_margarine

# Load the GSSvocab dataset
data("GSSvocab")
df2 <- GSSvocab
```


```{r}
#take a look at the data and summary
head(df)
summary(df)

```


# Part 1



Investigate  the  correlation between  margarine  consumption  and  divorce  rates  in  Maine.  Would  an  increase  in  the preference for margarine lead to skyrocketing divorce rates?

```{r}
# Visualize relationship
ggplot(df, aes(x = margarine_consumption_per_capita, y = divorce_rate_maine)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Relationship between margarine consumption and divorce rate",
       x = "Margarine Concumption",
       y = "Divorce Rate") +
  theme_minimal()
```


```{r}
#correlation test
cor.test(df$margarine_consumption_per_capita, df$divorce_rate_maine , method = c("pearson"))
```

```{r}
#making a linear model
lm(df$divorce_rate_maine ~ df$margarine_consumption_per_capita)
```

Investigate  the  correlation between  margarine  consumption  and  divorce  rates  in  Maine.  Would  an  increase  in  the preference for margarine lead to skyrocketing divorce rates?

*Answer:*
Based on visual inspection of the data as well as a correlation test (cor.test), the results indicate that there is a very strong positive correlation between margarine consumption per capita and divorce rates in Maine with a correlation coefficient equal to 0.9925585. This suggests that as margarine consumption increases, the divorce rate tends to increase as well. Even though, correlation is statistically significant (p-value = 1.33e-08), without the knowledge of the theoretical background of the data set we cannot assume any causality between the variables. The observed correlation could be influenced by other factors or variables that are not accounted for in this analysis. To make any causality conclusions it is essential to explore the context with further investigation to understand the present relationships better.



# Part 2



This dataset contains people’s scores on an English vocabulary test and includes demographic information. Filter for the year 1978 and remove rows with missing values (the function na.exclude()is one way to do this–check out the documentation!).

```{r}
head(df2)
summary(df2)
```

```{r}
# Filter for rows where year is 1978
df2_filtered <- df2 %>% filter(year == 1978)

head(df2_filtered)
summary(df2_filtered) #also checking for NAs
```
```{r}
# Remove rows with NAs using na.exclude
df2_filtered <- na.exclude(df2_filtered)

# Check for NAs with summary
summary(df2_filtered)
```


## 2.1


Is a person’s score on the vocabulary test (‘vocab’) significantly impacted by their level of education (‘educ’)?
Visualize the relationship in a plot and build a model. Briefly explain the results.


```{r}
# Visualize relationship
ggplot(df2_filtered, aes(x = educ, y = vocab)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Relationship between level of education and the vocabulary test score",
       x = "Education Level",
       y = "Vocabulary test score") +
  theme_minimal()

```
```{r}
#make a model
m1 <- lm(vocab~educ, df2_filtered)
summary(m1)
```

*Answer:*

This model indicates a positive and significant association between education level and vocabulary scores. The low p-value implies that the effect of education on vocabulary score is statistically significant. Nevertheless education alone explains only partly the variation in vocabulary, as the R-squared value of 0.288 suggests that about 28.8% of the variability in vocabulary scores can be explained by education level.


## 2.2


Whether a person is the native of an English-speaking country (‘nativeBorn’) could potentially have an impact on the size of their vocabulary. Visualize the relationship and add the predictor to the model. Briefly explain the results.

```{r}
#Visualize the relationship
ggplot(df2_filtered, aes(x = educ, y = vocab, color = nativeBorn)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  labs(
    title = "Relationship between native language and the vocabulary test score",
    x = "Education level",
    y = "Vocabulary test score",
    color = "Native Born",
    linetype = "Native Born"
  ) +
  scale_color_manual(values = c("blue", "red"), labels = c("No", "Yes")) +
  theme_minimal()

```

```{r}
#make a model and add a predictor
m2 <- lm(vocab ~ educ + nativeBorn, df2_filtered)
summary(m2)
```

*Answer:*

The estimate for nativeBornyes (0.65) suggests that being a native English speaker is associated with an additional 0.65 points in vocabulary score, holding education level consistent with the previous model. This effect is statistically significant (p = 0.00159).
The R-squared value slightly increased to 0.293, indicating that adding nativeBorn improved the model’s explanatory power, though the increase is small. The results suggest that higher education levels predict higher vocabulary scores and being a native of an English-speaking country adds a modest but significant increase in vocabulary scores.


## 2.3


Does  a  person’s  level  of  education  depend  on  whether they  are  a  native  of  the  country? Do you think it makes sense to add the relationship as an interaction term? Try creating the model and briefly explain the results. 

```{r}
#make a model with interaction
m3 <- lm(vocab ~ educ * nativeBorn, df2_filtered)
summary(m3)
```

*Answer:*

This model indicates that native-born status does not significantly impact baseline vocabulary scores when education is zero, and the interaction term between education and native-born status is also not significant (p = 0.663). This indicates that the relationship between education and vocabulary score is similar for both native and non-native English speakers. Adding the interaction term does not improve the model’s explanatory power, as reflected in the very small change in R-squared (remaining around 0.293). Thus, the simpler model without the interaction term is both sufficient and more interpretable, with education having a consistent positive effect on vocabulary regardless of native-born status.


## 2.4


Which model performs best? 

```{r}
#compare models
anova(m1, m2)
anova(m2, m3)
anova(m1, m3)
```
```{r}
# ANOVA to compare models
anova(m1, m2, m3)

# AIC and BIC for model comparison
AIC(m1, m2, m3)
BIC(m1, m2, m3)

```

*Answer:*

The ANOVA test indicates that Model 2 (with educ + nativeBorn) significantly improves the fit over Model 1 (with only educ as a preditor), with a p-value of 0.00159. However, adding the interaction term in Model 3 does not provide a significant improvement over Model 2 (p = 0.66344). Model 2 has the lowest AIC (6060.4) and BIC (6081.6) values, suggesting it has the best balance between model fit and complexity. Model 3 has slightly higher AIC and BIC values, indicating that the added complexity of the interaction term does not improve model performance. Overall, Model 2 (vocab ~ educ + nativeBorn) performs best, as it significantly improves fit over the baseline model (m1) and has the lowest AIC and BIC scores without unnecessary complexity.





